---
title: "Introduction to the dsge Package"
author: "Mustapha Wasseja Mohammed"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the dsge Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `dsge` package specifies, solves and estimates dynamic stochastic
general equilibrium (DSGE) models in R, without Dynare, MATLAB or Octave.
It covers:

- **Models**: linear models written with a formula interface
  (`dsge_model()`), nonlinear models written as equations
  (`dsgenl_model()`), and models imported from Dynare `.mod` files
  (`read_dynare()`).
- **Solutions**: first-order solutions, and second- and third-order
  perturbation for nonlinear models; perfect-foresight simulations and
  occasionally binding constraints (OccBin).
- **Estimation**: maximum likelihood with the Kalman filter, and Bayesian
  estimation by adaptive random-walk Metropolis-Hastings, sequential Monte
  Carlo or particle MCMC.
- **Analysis**: impulse responses, forecasts, conditional forecasts,
  variance and historical shock decompositions, identification
  diagnostics and optimal policy.

This vignette walks through the core workflow, from a one-equation model
to imported Dynare models.

## A Simple AR(1) Model

We start with the simplest possible DSGE model: a single observed variable
driven by a single autoregressive state variable.

The model is:
$$y_t = z_t$$
$$z_{t+1} = \rho z_t + \varepsilon_{t+1}$$

where $\varepsilon_t \sim N(0, \sigma^2)$.

### Simulate Data

```{r simulate}
set.seed(42)
true_rho <- 0.8
true_sd <- 1.0
n <- 200

z <- numeric(n)
for (i in 2:n) z[i] <- true_rho * z[i - 1] + true_sd * rnorm(1)
dat <- data.frame(y = z)
```

### Specify the Model

```{r specify}
library(dsge)

m <- dsge_model(
  obs(y ~ z),
  state(z ~ rho * z),
  start = list(rho = 0.5)
)

print(m)
```

### Solve at Known Parameters

Before estimating, we can solve the model at specific parameter values
to inspect the state-space solution.

```{r solve}
sol <- solve_dsge(m, params = c(rho = 0.8))
print(sol)
```

### Estimate the Model

```{r estimate}
fit <- estimate(m, data = dat)
summary(fit)
```

The estimated value of `rho` should be close to the true value of 0.8.

### Postestimation

```{r postest}
# Policy matrix (maps states to controls)
policy_matrix(fit, se = FALSE)

# Transition matrix (state dynamics)
transition_matrix(fit, se = FALSE)

# Stability check
stability(fit)
```

### Impulse-Response Functions

```{r irf, fig.height=3}
irfs <- irf(fit, periods = 20, se = FALSE)
plot(irfs)
```

The IRF shows the response of `y` to a one-standard-deviation shock
to the state variable `z`. The response decays geometrically at rate `rho`.

### Forecasting

```{r forecast, fig.height=3}
fc <- forecast(fit, horizon = 12)
plot(fc)
```

The forecast converges toward the unconditional mean of the series.

## A Three-Equation New Keynesian Model

A more realistic example is a simple New Keynesian model with three
observed/unobserved variables:

- $p_t$: inflation (observed)
- $x_t$: output gap (unobserved)
- $r_t$: interest rate (observed)

The linearized model is:

$$p_t = \beta E_t[p_{t+1}] + \kappa x_t$$
$$x_t = E_t[x_{t+1}] - (r_t - E_t[p_{t+1}] - g_t)$$
$$r_t = \psi p_t + u_t$$
$$u_{t+1} = \rho_u u_t + \varepsilon_{u,t+1}$$
$$g_{t+1} = \rho_g g_t + \varepsilon_{g,t+1}$$

### Model Specification

```{r nk-model}
nk <- dsge_model(
  obs(p   ~ beta * lead(p) + kappa * x),
  unobs(x ~ lead(x) - (r - lead(p) - g)),
  obs(r   ~ psi * p + u),
  state(u ~ rhou * u),
  state(g ~ rhog * g),
  fixed = list(beta = 0.96),
  start = list(kappa = 0.1, psi = 1.5, rhou = 0.7, rhog = 0.9)
)

print(nk)
```

This model has five equations, three control variables (two observed,
one unobserved), and two exogenous state variables. The discount factor
$\beta$ is constrained to 0.96.

### Simulate and Estimate

We simulate data from the model at known parameter values and then
estimate the parameters by maximum likelihood.

```{r nk-sim-est}
# Solve at true parameters to get state-space form
true_params <- c(beta = 0.96, kappa = 0.085, psi = 1.94,
                 rhou = 0.70, rhog = 0.95)
true_sd <- c(u = 2.3, g = 0.57)
sol <- solve_dsge(nk, params = true_params, shock_sd = true_sd)

# Simulate data
set.seed(123)
n <- 200
states <- matrix(0, n, 2)
colnames(states) <- c("u", "g")
for (t in 2:n) {
  states[t, "u"] <- 0.70 * states[t - 1, "u"] + 2.3 * rnorm(1)
  states[t, "g"] <- 0.95 * states[t - 1, "g"] + 0.57 * rnorm(1)
}
Z <- sol$D %*% sol$G
obs_data <- states %*% t(Z)
colnames(obs_data) <- c("p", "r")
nk_dat <- as.data.frame(obs_data)

# Estimate
nk_fit <- estimate(nk, data = nk_dat,
                   start = list(kappa = 0.1, psi = 1.5,
                                rhou = 0.7, rhog = 0.9),
                   control = list(maxit = 500))
summary(nk_fit)
```

### Impulse-Response Functions

```{r nk-irf, fig.height=5}
nk_irfs <- irf(nk_fit, periods = 20)
plot(nk_irfs)
```

The IRF panels show how inflation, interest rate, and the output gap
respond to monetary policy (u) and demand (g) shocks.

### Forecasting

```{r nk-forecast, fig.height=4}
nk_fc <- forecast(nk_fit, horizon = 12)
plot(nk_fc)
```

### Variance and Historical Decompositions

How much of each variable's variance does each shock explain? Without a
horizon, `variance_decomposition()` gives the unconditional shares; with
one, the forecast-error variance decomposition at those horizons.

```{r nk-vd, fig.height=3.5}
variance_decomposition(nk_fit, horizon = c(1, 4, 8))
plot(variance_decomposition(nk_fit))
```

`shock_decomposition()` splits each observed series, period by period, into
the contributions of the (smoothed) shocks:

```{r nk-hd, fig.height=5}
plot(shock_decomposition(nk_fit))
```

## Nonlinear DSGE Models

Nonlinear models are written as equations in strings, with `VAR(+1)` for
next-period values. They are solved by perturbation around the
deterministic steady state: to first order by default, and to second or
third order on request.

### A Simple RBC Model

```{r rbc-model}
rbc <- dsgenl_model(
  "1/C = beta / C(+1) * (alpha * exp(Z) * K^(alpha-1) + 1 - delta)",
  "K(+1) = exp(Z) * K^alpha - C + (1 - delta) * K",
  "Z(+1) = rho * Z",
  observed = "C",
  endo_state = "K",
  exo_state = "Z",
  fixed = list(alpha = 0.33, beta = 0.99, delta = 0.025),
  start = list(rho = 0.9),
  ss_guess = c(C = 2, K = 30, Z = 0)
)
print(rbc)
```

### Steady State

```{r rbc-ss}
params <- c(alpha = 0.33, beta = 0.99, delta = 0.025, rho = 0.9)
ss <- steady_state(rbc, params = params)
print(ss)
```

### Solve and Inspect

`solve_dsge()` automatically linearizes and solves:

```{r rbc-solve}
sol <- solve_dsge(rbc, params = params, shock_sd = c(Z = 0.01))
print(sol)
stability(sol)
```

### IRFs

```{r rbc-irf, fig.height=3}
rbc_irfs <- irf(sol, periods = 40, se = FALSE)
plot(rbc_irfs)
```

The IRF shows how consumption responds to a TFP shock, with the
response decaying as capital adjusts back to steady state.

### Higher-Order Solutions

With `order = 2` (or `3`), `solve_dsge()` adds the second-order terms of the
decision rules. The constant `g_ss` is the risk correction: how far
precautionary behaviour moves the controls away from the deterministic
steady state.

```{r rbc-order2}
sol2 <- solve_dsge(rbc, params = params, shock_sd = c(Z = 0.01), order = 2)
sol2$g_ss
```

Second-order responses depend on the size and sign of the shock. For a
large (five standard deviation) shock, the fall in consumption after a
negative shock is not the mirror image of the rise after a positive one:

```{r rbc-girf, fig.height=3}
up <- irf_2nd_order(sol2, shock = "Z", size = 5, periods = 40)
down <- irf_2nd_order(sol2, shock = "Z", size = -5, periods = 40)
c_up <- up$response[up$variable == "C"]
c_down <- down$response[down$variable == "C"]
round(c(impact_up = c_up[1], impact_down = c_down[1],
        asymmetry = c_up[1] + c_down[1]), 6)
plot(up, variables = "C")
```

## Importing Dynare Models

`read_dynare()` reads a Dynare `.mod` file (or model code given as text),
including its calibration, steady state, shocks, priors and estimation
settings, and returns a model that the functions above accept directly.

```{r dynare-import}
nk_mod <- read_dynare(text = "
  var y pi i u;
  varexo e_d e_m;
  parameters beta sigma kappa phi_pi phi_y rho_i rho_u;
  beta = 0.99; sigma = 1; kappa = 0.1;
  phi_pi = 1.5; phi_y = 0.125; rho_i = 0.7; rho_u = 0.8;
  model(linear);
    y = y(+1) - 1/sigma * (i - pi(+1)) + u;
    pi = beta * pi(+1) + kappa * y;
    i = rho_i * i(-1) + (1 - rho_i) * (phi_pi * pi + phi_y * y) + e_m;
    u = rho_u * u(-1) + e_d;
  end;
  shocks;
    var e_d; stderr 0.01;
    var e_m; stderr 0.0025;
  end;
")
nk_sol <- solve_dsge(nk_mod)
```

```{r dynare-irf, fig.height=5}
plot(irf(nk_sol, periods = 20, se = FALSE), response = c("y", "pi", "i"))
```

Files with `perfect_foresight_setup` and `perfect_foresight_solver` (or
`simul`) are simulated with `simulate_perfect_foresight()`. Here capital and
consumption move to a new steady state after a permanent 10% rise in
productivity:

```{r dynare-pf, fig.height=3}
pf_mod <- read_dynare(text = "
  var c k;
  varexo a;
  parameters alpha beta delta;
  alpha = 0.33; beta = 0.99; delta = 0.025;
  model;
    1/c = beta/c(+1) * (alpha * exp(a(+1)) * k^(alpha - 1) + 1 - delta);
    k = exp(a) * k(-1)^alpha + (1 - delta) * k(-1) - c;
  end;
  steady_state_model;
    k = ((1/beta - 1 + delta)/(alpha*exp(a)))^(1/(alpha - 1));
    c = exp(a)*k^alpha - delta*k;
  end;
  initval; a = 0; end;
  steady;
  endval; a = 0.1; end;
  steady;
  perfect_foresight_setup(periods = 200);
  perfect_foresight_solver;
")
plot(simulate_perfect_foresight(pf_mod))
```

Occasionally binding constraints declared with Dynare's `occbin_constraints`
are simulated with `simulate_occbin()`, and optimal-policy commands
(`ramsey_model`, `discretionary_policy`, `osr`) are translated as well; see
`?read_dynare`.

## Bayesian Estimation

`bayes_dsge()` estimates linear and nonlinear models by adaptive
random-walk Metropolis-Hastings (RWMH), with parallel chains. For
nonlinear models, the steady state is re-solved and the model re-linearized
at each candidate parameter draw. Alternatives are `bayes_smc()` (tempered
sequential Monte Carlo) and `bayes_particle()` (particle marginal
Metropolis-Hastings with a bootstrap particle filter), and
`bayes_factor()` compares models by their marginal likelihoods.

### Specifying Priors

Use `prior()` to define prior distributions for structural parameters:

```{r bayes-priors, eval=FALSE}
my_priors <- list(
  beta  = prior("beta", shape1 = 95, shape2 = 5),
  kappa = prior("beta", shape1 = 30, shape2 = 70),
  psi   = prior("gamma", shape = 20, rate = 13.3),
  rhou  = prior("beta", shape1 = 70, shape2 = 20),
  rhog  = prior("beta", shape1 = 70, shape2 = 20)
  # shock SDs default to inv_gamma(0.01, 0.01)
)
```

Supported distributions: `normal`, `beta`, `gamma`, `uniform`, `inv_gamma`
and `inv_gamma1` (Dynare's type-1 inverse gamma).

### Running the Sampler

```{r bayes-estimate, eval=FALSE}
fit_bayes <- bayes_dsge(nk, data = nk_dat, priors = my_priors,
                        chains = 2, iter = 10000, warmup = 5000,
                        seed = 42)

summary(fit_bayes)  # posterior table with ESS, R-hat, MCSE
```

### Posterior Diagnostics and IRFs

```{r bayes-diag, eval=FALSE}
# MCMC diagnostics
plot(fit_bayes, type = "trace")            # trace plots (all chains)
plot(fit_bayes, type = "density")          # posterior density + prior overlay
plot(fit_bayes, type = "prior_posterior")   # dedicated prior-vs-posterior comparison
plot(fit_bayes, type = "running_mean")     # cumulative mean convergence
plot(fit_bayes, type = "acf")              # autocorrelation diagnostics
plot(fit_bayes, type = "pairs")            # pairwise scatter + correlations
plot(fit_bayes, type = "all")              # combined trace + density panel

# Parameter selection (works with all types above)
plot(fit_bayes, type = "trace", pars = c("kappa", "psi"))
plot(fit_bayes, type = "density", pars = "rhou")

# Posterior IRFs with credible bands
plot(fit_bayes, type = "irf", periods = 12, n_draws = 500)
# Or equivalently:
bayes_irfs <- irf(fit_bayes, periods = 12, n_draws = 500)
plot(bayes_irfs)
```

**Available plot types for `dsge_bayes` objects:**

| Type | Description |
|------|-------------|
| `"trace"` | MCMC trace plots (all chains overlaid) |
| `"density"` | Posterior density with prior overlay |
| `"prior_posterior"` | Dedicated prior-vs-posterior comparison |
| `"running_mean"` | Cumulative posterior mean by iteration |
| `"acf"` | Autocorrelation function plots |
| `"pairs"` | Pairwise scatter, correlations, histograms |
| `"all"` | Combined trace + density side by side |
| `"irf"` | Posterior IRFs with credible bands |

The `pars` argument selects specific parameters (e.g.,
`pars = c("kappa", "psi")`). Forecast plotting is not yet available
for Bayesian fits.

**Note:** `bayes_dsge()` evaluates the likelihood of nonlinear models with
the Kalman filter on the first-order solution around parameter-specific
steady states.

For examples:

- Linear Bayesian on real data: `inst/examples/bayesian_nk_fred.R`
- Nonlinear Bayesian RBC: `inst/examples/bayesian_rbc_nonlinear.R`
- Nonlinear Bayesian NK: `inst/examples/bayesian_nk_nonlinear.R`
- RBC with capital accumulation: `inst/examples/bayesian_rbc_capital_demo.R`
  (quick demo, ~2.5 min) or `bayesian_rbc_capital_full.R` (full, ~9 min)

## ggplot2 Graphics

The `plot()` methods above use base graphics, so they work with no extra
packages. If you use **ggplot2**, `autoplot()` draws the same results as
ggplot objects: IRFs (including second order), forecasts, variance and
historical decompositions, smoothed states, perfect-foresight and OccBin
paths, and posterior traces and densities. ggplot2 is optional; the
chunks below run only when it is installed.

```{r gg-setup, include = FALSE}
has_gg <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r gg-irf, eval = has_gg, fig.height = 5}
library(ggplot2)
autoplot(nk_irfs)
```

```{r gg-forecast, eval = has_gg, fig.height = 4}
autoplot(nk_fc)
```

Because the result is an ordinary ggplot, you can add layers, titles or
facets, or replace the theme:

```{r gg-fevd, eval = has_gg, fig.height = 3.5}
autoplot(variance_decomposition(nk_fit, horizon = c(1, 4, 8, 20))) +
  labs(subtitle = "Share of the forecast-error variance due to each shock")
```

```{r gg-hd, eval = has_gg, fig.height = 5}
autoplot(shock_decomposition(nk_fit))
```

`theme_dsge()`, `scale_colour_dsge()` and `scale_fill_dsge()` give your
own ggplots the same look. Here the second-order responses of consumption
and capital from the RBC section are drawn with ggplot2 directly, one panel
each:

```{r gg-theme, eval = has_gg, fig.height = 3}
rbc_up <- up[up$variable %in% c("C", "K"), ]
ggplot(rbc_up, aes(period, response, colour = variable)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_line(linewidth = 0.9, show.legend = FALSE) +
  facet_wrap(~ variable, scales = "free_y") +
  scale_colour_dsge() +
  labs(title = "Second-order responses to a 5 s.d. technology shock",
       x = "Period", y = NULL) +
  theme_dsge()
```

For a Bayesian fit, `autoplot(fit_bayes)` draws the MCMC traces and
`autoplot(fit_bayes, type = "density")` the posterior densities with the
priors overlaid.

## Other Tools

- **Forecasting**: `conditional_forecast()` fixes the future path of some
  variables; `predict()`, `fitted()` and `residuals()` work on fitted models.
- **Filtering and smoothing**: `smooth_states()`, `smooth_shocks()`,
  `particle_filter()`, and `kalman_filter_skewed()` for skew-normal shocks.
- **Optimal policy**: `ramsey_policy()`, `discretionary_policy()` and
  optimal simple rules with `osr()`.
- **Other estimators**: `irf_match()`, `gmm_estimate()`, `smm_estimate()`,
  DSGE-VAR (`bayes_dsge_var()`, `bayes_dsge_var_mh()`) and endogenous priors.
- **Diagnostics**: `check_identification()`, `parameter_sensitivity()`,
  `global_sensitivity()` and `robust_vcov()`.
- **Nonlinear dynamics**: `perfect_foresight_nonlinear()`, `extended_path()`
  and `simulate_occbin()`.
- **Reporting**: `model_latex()` writes the model equations in LaTeX.

## Notes

- For linear models, `lead()` or `E()` marks forward expectations.
- For nonlinear models, `VAR(+1)` denotes the next-period value.
- In linear models written with `dsge_model()`, the number of exogenous
  state variables (with shocks) must equal the number of observed
  variables. Nonlinear models and imported Dynare models may have more
  shocks than observed variables, and measurement errors.
- Nonlinear models are solved by perturbation around the deterministic
  steady state. First-order solutions use the same solver and estimation
  pipeline as linear models.
